---
title: "Getting started with arsbridge"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with arsbridge}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE     # Vignette is documentation; calls are illustrative.
)
```

## What `arsbridge` does

`arsbridge` reads a lead programmer's already-annotated TLF shell Word
document plus the study's ADaM specification, and produces a CDISC
Analysis Results Standard (ARS) v1.0 ARM-TS JSON file. That JSON is
consumed downstream by
[`siera::readARS()`](https://pharmaverse.github.io/siera/) to generate
the R analysis code that produces the actual TLFs.

**Core principle:** the package extracts and converts -- it does not
invent. The shell is read by a deterministic regex detector *and* an LLM
primary reader together, to capture as many annotation variants as
possible; but every LLM-proposed variable is gated against the ADaM spec,
so a variable absent from the spec is rejected and logged, never shipped.
Every variable in the ARS output traces back to a real annotation
grounded in the spec. For the full reading process see
`vignette("reading-engine")`.

## Prerequisites

```{r}
install.packages("arsbridge")            # or devtools::install_github(...)

# One-time API key setup (interactive prompt)
arsbridge::set_anthropic_key()
arsbridge::check_anthropic_key()         # confirm
```

`arsbridge` calls Claude once per TLF section for light semantic
enrichment (analysis type, method name, row role). All variable names
come from the shell annotations -- never the LLM.

## The fastest path: `spec_to_ars_example()`

The package ships with a small bundled training example, so you can run
the entire pipeline before you own a study:

```{r}
library(arsbridge)

res <- spec_to_ars_example()
# ARS JSON and validation report land in tempdir(); paths are in res.
res$n_tlfs        # 40
res$n_analyses    # ~226
res$n_warnings    # ~29
```

Takes about six minutes (40 LLM calls).

## Inspecting the result

```{r}
# Top-level shape of the ARS ReportingEvent
str(res$reporting_event, max.level = 1)

# Validation findings -- one row per annotation
table(res$validation$status)
subset(res$validation, status %in% c("WARN", "FAIL"))

# One output, end to end
out <- Filter(function(o) o$name == "T-14-1-2",
              res$reporting_event$outputs)[[1]]
length(out$referencedAnalysisIds)
```

## From ARS to a formatted clinical table

The ARS JSON plus the study ADaM data closes the loop: `ars_to_ard()`
executes the analyses into a tidy ARD (`{cards}` format), and
`ars_render_tlf()` formats any output into a publication-ready GT table
-- no manual formatting step.

```{r}
# Unzip the bundled simulated ADaM data
adam_dir <- file.path(tempdir(), "ADaM")
unzip(arsbridge_example("ADaM.zip"), exdir = adam_dir)

# Execute the ARS specification into a tidy ARD
ard <- ars_to_ard(
  ars_path = "outputs/reporting_event.json",
  adam_dir = adam_dir
)

# Render one output straight to a formatted GT clinical table
gt_table <- ars_render_tlf(
  ars_path  = "outputs/reporting_event.json",
  ard       = ard,
  output_id = "T_14_1_1"      # Subject Disposition
)

gt_table
```

`ars_render_tlf()` auto-detects the treatment column, row groups, and row
labels from the ARD; rescales `{cards}` proportions to percentages; lays
continuous summaries out as `Mean (SD)` / `Median` / `(Min, Max)` rows;
and carries ARS footnotes through as GT source notes.

To build the `tfrmt` spec without rendering -- so you can tweak it before
printing -- use `ars_to_tfrmt()`. To get one spec per output at once, use
`ars_to_tfrmt_list()`:

```{r}
specs <- ars_to_tfrmt_list("outputs/reporting_event.json", ard)
names(specs)

all_tables <- lapply(
  names(specs),
  function(oid) ars_render_tlf("outputs/reporting_event.json", ard, oid)
)
```

## Using your own files

```{r}
res <- spec_to_ars(
  shell_path     = "inputs/annotated_shell.docx",
  adam_spec_path = "inputs/adam_spec.xlsx",        # or define.xml
  output_path    = "outputs/reporting_event.json",
  report_path    = "outputs/spec_validation_report.xlsx",
  study_id       = "ABC-123",
  study_name     = "ABC-123 Phase 3"
)
```

`adam_spec_path` accepts either `.xml` (ADaM define.xml -- preferred
when available) or `.xlsx`/`.xls` (ADaM spec Excel -- used during
development before define.xml exists). The SDTM spec is NOT a valid
input -- TLF annotations reference ADaM variables.

## The bundled example, file by file

```{r}
arsbridge_example()                              # list bundle contents
arsbridge_example("annotated_shell.docx")        # absolute path
arsbridge_example("adam_spec.xlsx")
arsbridge_example("ADaM.zip")                    # for downstream siera
```

| File | Purpose |
|---|---|
| `annotated_shell.docx` | The TLF shells with red ADaM annotations |
| `adam_spec.xlsx` | The ADaM variable spec |
| `ADaM.zip` | 60-subject simulated ADaM data -- not consumed by arsbridge, but used by the downstream siera runner |

## What the parser actually detects

`arsbridge` uses a four-layer detection hierarchy and stops at the
highest-confidence layer that produces a result:

1. **Colour-based** -- red `#C00000` runs that match an ADaM pattern. HIGH confidence.
2. **Formatting-based** -- bold / italic / underline runs that match. MEDIUM confidence.
3. **Plain-text pattern** -- `DATASET.VARIABLE` strings recognised by regex alone, no formatting needed. HIGH confidence (the structure is unique).
4. **LLM fallback** -- only for cells where 1-3 produce nothing and context implies an annotation should exist. LOW confidence.

In practice, the bundled `annotated_shell.docx` triggers Layer 1 for all
130 stub-row annotations -- the convention is consistent.

## Listing column-header detection

Listings differ from tables: the variable lives in the column header,
not the stub. Cells like

```
Subject ID
USUBJID
```

get split -- "Subject ID" is the label, `USUBJID` becomes the
annotation. The dataset prefix is resolved by combining a universal
ADSL list (USUBJID, ARM, TRT01A, etc. always live in ADSL) with the
TLF's source dataset from the `Source: ADAE` line below the table.

## Closing the loop -- downstream siera

The package ships with a standalone runner at `siera_workflow/` in the
project repository (not bundled in the package itself). It takes the
ARS JSON arsbridge produces, runs it through `siera::readARS()` to
generate R analysis code, and executes that code against the simulated
`ADaM.zip`. See `siera_workflow/README.md` for the full chain.

## What arsbridge will NOT do

- Invent variable names from shell row labels or titles
- Read or process ADaM data files (XPT, sas7bdat, CSV) -- siera's job
- Write Word documents (no shell annotation -- arsbridge only reads)
- Generate analysis R code -- siera's job

## Where to look when something goes wrong

- `res$validation` and `outputs/spec_validation_report.xlsx` -- per-annotation PASS/WARN/FAIL
- `cli::cli_alert_warning` lines during the run -- usually a TLF with an unexpected structure
- Bundled test fixtures at `tests/testthat/fixtures/` -- minimal known-good shells to compare against
