---
title: "Getting started with arsbridge"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with arsbridge}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE     # Vignette is documentation; calls are illustrative.
)
```

## What arsbridge does

`arsbridge` reads a lead programmer's annotated TLF shell Word document
together with the study's ADaM specification, and produces a CDISC
Analysis Results Standard (ARS) v1.0 JSON file. That JSON drives the
rest of the pipeline: it executes against real ADaM datasets using
`{cards}` to produce a tidy ARD, then renders to a formatted GT clinical
table with no manual formatting step.

**The core principle:** arsbridge extracts and converts. It does not invent.
The shell is read by a deterministic regex detector *and* an LLM primary
reader together, to capture as many annotation variants as possible. Every
LLM-proposed variable is then checked against the ADaM spec, so a variable
absent from the spec is rejected and logged, never shipped. Every variable
in the ARS output traces back to a real annotation grounded in the spec.

For the full reading-engine detail, see `vignette("reading-engine")`.

---

## Prerequisites

```{r}
# Install from GitHub
devtools::install_github("tavakohr/arsbridge")

# One-time API key setup (interactive prompt; hides input when {askpass} is installed)
arsbridge::set_anthropic_key()
arsbridge::check_anthropic_key()   # confirm the key loaded
```

`arsbridge` calls Claude once per TLF section for light semantic enrichment:
analysis type, method name, and row role. All variable names come from the
shell annotations, never from the LLM.

---

## The fastest path: `spec_to_ars_example()`

The package ships with a small bundled training example so you can run the
entire pipeline before you own a study. This is the best way to get a feel
for the output and the validation report.

```{r}
library(arsbridge)

res <- spec_to_ars_example()
# ARS JSON and validation report land in tempdir(); paths are in res.
res$n_tlfs        # 40 TLF shells
res$n_analyses    # ~226 analyses
res$n_warnings    # ~29 warnings to review
```

Takes about six minutes (40 LLM calls, one per TLF section).

---

## Inspecting the result

Once `spec_to_ars_example()` finishes, you have a structured
`reporting_event` object and a row-level validation report.

```{r}
# Top-level shape of the ARS ReportingEvent
str(res$reporting_event, max.level = 1)

# Validation findings: one row per annotation
table(res$validation$status)
subset(res$validation, status %in% c("WARN", "FAIL"))

# Drill into one output
out <- Filter(function(o) o$name == "T-14-1-2",
              res$reporting_event$outputs)[[1]]
length(out$referencedAnalysisIds)
```

The **PASS / WARN / FAIL** stamps tell you:

| Status | Meaning |
|---|---|
| **PASS** | Variable found in ADaM spec; annotation is clean. |
| **WARN** | Regex and LLM disagree; human review recommended. |
| **FAIL** | Variable not found in ADaM spec; will be rejected as a blocker. |

---

## From ARS to a formatted clinical table

With a valid ARS JSON in hand, two more functions close the loop.

`ars_to_ard()` executes the ARS against ADaM datasets and returns a tidy
ARD in `{cards}` format. `ars_render_tlf()` then formats any output from
that ARD into a publication-ready GT table, with no manual formatting step.

```{r}
# Unzip the bundled simulated ADaM data
adam_dir <- file.path(tempdir(), "ADaM")
unzip(arsbridge_example("ADaM.zip"), exdir = adam_dir)

# Execute the ARS specification into a tidy ARD
ard <- ars_to_ard(
  ars_path = res$ars_path,
  adam_dir = adam_dir
)

# Render the Subject Disposition table (T_14_1_1) as a GT clinical table
gt_table <- ars_render_tlf(
  ars_path  = res$ars_path,
  ard       = ard,
  output_id = "T_14_1_1"
)

gt_table
```

`ars_render_tlf()` auto-detects the treatment column, row groups, and row
labels from the ARD; rescales `{cards}` proportions to percentages; lays
continuous summaries out as `Mean (SD)` / `Median` / `(Min, Max)` rows;
and carries ARS footnotes through as GT source notes.

To inspect or customise the underlying `{tfrmt}` spec before printing:

```{r}
# One output: get the tfrmt spec
spec <- ars_to_tfrmt(res$ars_path, ard, "T_14_1_1")

# All outputs at once
specs <- ars_to_tfrmt_list(res$ars_path, ard)
names(specs)

all_tables <- lapply(
  names(specs),
  function(oid) ars_render_tlf(res$ars_path, ard, oid)
)
```

---

## Using your own study files

```{r}
res <- spec_to_ars(
  shell_path     = "inputs/annotated_shell.docx",
  adam_spec_path = "inputs/adam_spec.xlsx",        # or define.xml
  output_path    = "outputs/reporting_event.json",
  report_path    = "outputs/spec_validation_report.xlsx",
  study_id       = "ABC-123",
  study_name     = "ABC-123 Phase 3"
)
```

`adam_spec_path` accepts either `.xml` (ADaM `define.xml`, preferred when
available) or `.xlsx` / `.xls` (ADaM spec Excel, used during development
before `define.xml` exists). The SDTM spec is **not** a valid input: TLF
annotations reference ADaM variables.

---

## The bundled example, file by file

```{r}
arsbridge_example()                              # list all bundle contents
arsbridge_example("annotated_shell.docx")        # absolute path to a file
arsbridge_example("adam_spec.xlsx")
arsbridge_example("ADaM.zip")
```

| File | Purpose |
|---|---|
| `annotated_shell.docx` | TLF shells with red ADaM annotations (the input) |
| `adam_spec.xlsx` | ADaM variable spec (the grounding truth) |
| `ADaM.zip` | 60-subject simulated ADaM datasets (for `ars_to_ard()`) |

---

## What the shell parser actually detects

arsbridge uses a four-layer detection hierarchy and stops at the
highest-confidence layer that produces a result for each cell:

| Layer | Trigger | Confidence |
|---|---|---|
| 1. Colour-based | Red `#C00000` runs matching an ADaM pattern | HIGH |
| 2. Formatting-based | Bold / italic / underline runs matching a pattern | MEDIUM |
| 3. Plain-text pattern | `DATASET.VARIABLE` strings identified by regex alone | HIGH |
| 4. LLM pass | Cells where layers 1-3 produce nothing and context implies an annotation | Validated against spec |

In practice, the bundled `annotated_shell.docx` triggers Layer 1 for all
130 stub-row annotations. When shells use non-standard conventions, the LLM
pass catches what the regex misses.

---

## Listing column-header detection

Listings differ from summary tables: the variable lives in the column
header, not the stub. A cell like:

```
Subject ID
USUBJID
```

gets split automatically. "Subject ID" becomes the display label and
`USUBJID` becomes the annotation. The dataset prefix is resolved by
combining a universal ADSL list (variables like `USUBJID`, `ARM`, `TRT01A`
always live in ADSL) with the TLF's declared source dataset from a
`Source: ADAE` line below the table.

---

## Partial tables and the manual-fill round-trip

Some tables mix statistics arsbridge can compute (counts, percentages,
summary statistics) with ones it cannot (a Cochran-Mantel-Haenszel p-value,
an exact confidence interval, an NRI-imputed responder rate). arsbridge
does not fabricate the latter or drop the whole table. Instead it **reserves
a keyed cell** for each: the ARD carries a `manual_pending` stub row with
`stat = NA`, tied to the same output, analysis, and method as a real result.

List what is outstanding:

```r
ard <- ars_to_ard(ars_path, "inputs/ADaM")
ars_manual_worklist(ard)   # one row per reserved cell
```

When rendered, computed cells appear with their values and each reserved
cell shows a loud `[‡ manual]` marker keyed to a footnote. No blank cells,
no `NA`, no misleading zeros.

To fill a reserved cell, compute it with a validated analysis script, then
write the result back into the ARD row:

```r
i <- which(ard$result_status == "manual_pending")[1]
ard$stat[[i]]         <- 0.012
ard$result_status[i]  <- "manual_filled"
ard$value_source[i]   <- "manual"
ard$derivation_ref[i] <- "programs/cmh_t1421.R"
```

`derivation_ref` is the audit trail: not "someone typed 0.012" but "value
from `cmh_t1421.R`." Before rendering, arsbridge checks every manual fill:

```r
ars_validate_manual_fills(ard)   # zero rows = every fill is traceable
```

`ars_render_all()` runs this automatically and blocks any untraceable value
before it reaches the final document. The design and full rationale are in
`adr/0002-partial-results-traceability.md`.

---

## What arsbridge will NOT do

- Invent variable names from shell row labels or titles
- Write Word documents (it only reads shells, not annotates them)
- Generate R analysis code (that is `{siera}`'s job)
- Process SAS data files directly without conversion to `.xpt` or `.csv`

---

## Where to look when something goes wrong

| What to check | Where to look |
|---|---|
| Per-annotation PASS / WARN / FAIL | `res$validation` and `outputs/spec_validation_report.xlsx` |
| Runtime warnings during extraction | `cli::cli_alert_warning` lines printed during `spec_to_ars()` |
| Known-good minimal shells | `tests/testthat/fixtures/` in the package source |
| Blocker-only view | `arsbridge::ars_blockers(res$diagnostics)` |
| Everything recorded | `arsbridge::ars_diagnostics(res$diagnostics)` |
