---
title: "Using arsbridge without API access"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using arsbridge without API access}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE     # Vignette is documentation; calls are illustrative.
)
```

## Three tiers, one required

`arsbridge` has an LLM in the loop for two jobs: reading variable annotations
in shell layouts a regex was never designed for, and classifying each table
(analysis type, grouping columns, method). Neither is required. The reading
engine has three tiers and only the first is mandatory:

| Tier | You supply | Accuracy |
|---|---|---|
| **Deterministic** | shell + spec | Regex + keyword heuristics |
| **Supplement** | + a JSON file from a chat assistant | Near-LLM, no API call |
| **LLM** | + an API key | Full |

A missing key and a missing supplement never stop the run. With neither,
`spec_to_ars()` completes on regex plus heuristics and records the fact:

```{r}
res <- spec_to_ars("shell.docx", "adam_spec.xlsx")
res$extraction_mode
#> "deterministic"
```

Standard shells (variables in red text, `Label -> DATASET.VAR` lines,
bracketed `DATASET.VARIABLE` after a label) still bind correctly with no LLM
at all. What degrades in deterministic mode is variant annotation layouts,
multi-level groupings, the Total column, and analysis-type classification for
unusual tables — each recorded as a `WARN` in the diagnostics sheet.

## The supplement tier: use a chat assistant by hand

Many regulated or enterprise environments block LLM APIs but allow an
interactive chat assistant (GitHub Copilot, ChatGPT, an internal portal)
where you can upload files. The supplement tier turns that into near-LLM
accuracy with zero API calls: you ferry the work through the chat UI, and the
package consumes the assistant's reply as an auxiliary file.

### Step 1 — get the instruction file

```{r}
ars_copilot_instructions()
```

This writes two files into your working directory — the instruction file
`arsbridge_copilot_instructions.md` and the JSON Schema
`arsbridge_supplement_v3.schema.json` — and prints the workflow. The
instruction file is static and versioned: it tells the assistant exactly what
to read and the exact JSON shape (format version 3) to reply with, so the
package can validate the answer. For a large or complex shell where a single
pass misses items, use the two-session workflow instead:

```{r}
ars_copilot_instructions(workflow = "two_phase")
```

which writes a Phase-1 (evidence blueprint) and a Phase-2 (semantic
construction + repair) instruction file. Phase 1 returns
`tlf_extraction_blueprints.json`; you feed that back in session 2 to produce
the supplement.

### Typed conditions (format v3)

The single most important thing to know about the format: **every filter,
population, and column condition is a typed object, not a string.** A row that
displays "Completed" is expressed as

```json
{"rowLabel": "Completed", "variable": {"dataset": "ADSL", "variable": "EOSSTT"},
 "whereClause": {"condition": {"dataset": "ADSL", "variable": "EOSSTT",
                               "comparator": "EQ", "value": ["COMPLETED"]}}}
```

not as `"where": "EOSSTT='COMPLETED'"`. This is what eliminates the old
string-parsing failures (double `=`, smart quotes, `OR` splitting): the
assistant hands the package a parsed tree, so there is nothing left to
mis-parse. The instruction file and the shipped JSON Schema spell the shape out
in full, with a worked example.

### Step 2 — upload to the assistant

Upload **four files together** to your chat assistant:

* `arsbridge_copilot_instructions.md` (from step 1),
* `arsbridge_supplement_v3.schema.json` (so the assistant can self-check),
* your annotated TLF shell `.docx`,
* your ADaM specification `.xlsx`.

Save the assistant's reply — one JSON block — as `supplement.json`.

### Step 3 — pre-flight (optional but recommended)

```{r}
ars_validate_supplement("supplement.json", "adam_spec.xlsx")
```

This checks the format version, every typed condition, the analysis-type and
method enums, and — when you pass the spec — that every referenced variable
exists in your ADaM spec. Any `FAIL` line is worded so you can paste it back to
the assistant, e.g. *"regenerate: ADSL.WEIGHTBL is not in the ADaM spec"*; when
there are FAILs the result also carries a `repair_prompt` attribute bundling
them into one paste-ready block. When `jsonvalidate` is installed, the file is
additionally checked against the shipped JSON Schema.

### Step 4 — run

```{r}
res <- spec_to_ars(
  shell_path     = "shell.docx",
  adam_spec_path = "adam_spec.xlsx",
  supplement     = "supplement.json"
)
res$extraction_mode
#> "supplement"
```

## What the supplement can and cannot do

The supplement is an **assistant, not an override**. Two guarantees make it
safe to trust a draft it produced:

* **The regex wins (by default).** A supplement binding lands only on a row the
  deterministic pass left unannotated. If the shell already annotated a row,
  the shell's annotation stands and any disagreement is logged as a `WARN`
  for you to review — the assistant can never silently rewrite an authored
  annotation. When you *do* trust the assistant's reading more than the regex
  (a shell the deterministic pass extracts poorly), pass
  `spec_to_ars(supplement_trust = "prefer_supplement")`: a validated,
  spec-gated supplement value then overrides the regex on a conflict, with a
  `WARN` recording both and the shell's original kept as a secondary analysis.
  The spec gate still applies, so a hallucinated variable is rejected in either
  mode.
* **The spec is the gate.** Every `DATASET.VARIABLE` the supplement proposes
  is checked against your ADaM spec. One that is not in the spec is dropped
  and logged as a blocking `FAIL`, exactly as a live LLM's proposal would be.
  The assistant cannot invent a variable into your output.

* **It confirms the table set.** The assistant reads the whole shell
  independently and lists every output with its `title`, so `spec_to_ars()`
  can compare that inventory against what it parsed. A supplement entry with no
  matching table, a table the supplement never mentions, or a title that
  disagrees between the two is logged as a `WARN` — this is the one tier that
  can catch a table arsbridge parsed differently or missed, because the live
  LLM only ever sees sections the parser already found. If the shell heading
  carried no title but the supplement has one, the parsed section adopts it.

Statistic sub-rows (`Mean (SD)`, `Median`, `Q1, Q3`, `Min, Max`) are never
bound from the supplement: they belong to the analysis row above them, and
the layout-driven renderer fills them from that parent.

Everything downstream is unchanged — the ARS JSON is built, validated, and
emitted as per-TLF `{cards}` scripts the same way in every tier. The JSON's
`_meta.extraction_mode` field records which tier produced it, and
`_meta.requires_human_review` stays `TRUE`: whichever tier you used, a
qualified programmer must review the draft.

## Data governance

Uploading your shell and spec to a chat assistant transmits their **text** —
TLF titles, stub labels, footnotes, and the variable *names* from your spec —
to that provider. It does **not** transmit any patient data: shells are mock
by definition, and `spec_to_ars()` never reads ADaM datasets at this stage
(data are only touched later, locally, by `ars_to_ard()`). Even so, shell
wording and a study's variable inventory can be sponsor-confidential —
**confirm your organisation's policy before uploading.**

## Troubleshooting

| Symptom | Cause | Fix |
|---|---|---|
| `Supplement file is not valid JSON` | assistant wrapped the reply in prose, or used smart quotes | the reader strips a ```` ```json ```` fence and smart double-quotes automatically; if it still fails, ask the assistant to "reply with only one fenced strict-JSON block" |
| `Supplement version mismatch` | a v2 file, or one made with a different package version | v3 conditions are typed objects, not strings — regenerate with `ars_copilot_instructions()` from the installed version |
| `invalid whereClause` / `not a where-clause` (`FAIL`) | a condition was written as a string, or with a bad comparator | use the typed shape `{"condition": {"dataset", "variable", "comparator", "value": [...]}}`; paste the `repair_prompt` back to the assistant |
| `analysis label 'X' matched no stub row` (`WARN`) | the `rowLabel` is not the stub text verbatim | copy `rowLabel` exactly from the shell, footnote markers and all |
| `is not in the ADaM spec` (`FAIL`) | assistant proposed a variable your spec does not have | fix the analysis or add the variable to the spec |
| `Anchor mismatch` (`WARN`) | the supplement's first row / row count disagrees with the parsed table | confirm the assistant described the same table arsbridge parsed |
| `entry 'X' matches no TLF in the shell` (`WARN`) | TLF numbering typo in the reply | key each TLF by its shell heading number (e.g. `"14.1.1"`) |
| a TLF ran on heuristics (`WARN`) | that TLF is missing from the supplement | add it to `tlfs` and rerun |
