---
title: "Using arsbridge without API access"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using arsbridge without API access}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE     # Vignette is documentation; calls are illustrative.
)
```

## Three tiers, one required

Two jobs benefit from a language model: reading variable annotations in shell
layouts a regex was never designed for, and classifying each table (analysis
type, grouping columns, method). Neither requires live API access. The reading
engine has three tiers:

| Tier | You supply | Role |
|---|---|---|
| **Deterministic** | shell + spec | The baseline. Always runs; a supported mode on its own. |
| **Supplement** | + a reviewed JSON file from a chat assistant | **The preferred completion path.** Offline: no API key, no network, no `ellmer`. |
| **Live LLM** | + an API key and the optional `ellmer` package | An optional convenience for sites that permit it. |

The supplement tier is not a workaround for blocked environments -- it is the
intended production path. A reviewed JSON file is auditable, reproducible and
reviewable before it is applied, which a live API call is not, and it keeps the
model's contribution inside whatever environment your organisation already
trusts. A supplement also takes precedence: supplying one makes no live call
whatever `use_llm` says.

A missing key and a missing supplement never stop the run. With neither,
`spec_to_ars()` completes on regex plus heuristics and records the fact:

```{r}
res <- spec_to_ars("shell.docx", "adam_spec.xlsx")
res$extraction_mode
#> "deterministic"
```

Standard shells (variables in red text, `Label -> DATASET.VAR` lines,
bracketed `DATASET.VARIABLE` after a label) still bind correctly with no LLM
at all. What degrades in deterministic mode is variant annotation layouts,
multi-level groupings, the Total column, and analysis-type classification for
unusual tables — each recorded as a `WARN` in the diagnostics sheet.

## The supplement tier: use a chat assistant by hand

Most regulated and enterprise environments already provide an interactive chat
assistant (GitHub Copilot, ChatGPT, an internal portal) that can read uploaded
files, whether or not they permit outbound API calls. The supplement tier is
built around that: you ferry the work through the chat UI inside your own
environment, review the reply, and the package consumes it as an auxiliary
file. Nothing leaves the boundary your organisation already sanctions, and the
answer is a file you can read, diff and keep.

### Step 1 — get the instruction file

```{r}
ars_copilot_instructions()
```

This writes two files into your working directory — the instruction file
`arsbridge_copilot_instructions.md` and the JSON Schema
`arsbridge_supplement_v4.schema.json` — and prints the workflow. The
instruction file is static and versioned: it tells the assistant exactly what
to read and the exact JSON shape (format version 4) to reply with, so the
package can validate the answer. For a large or complex shell where a single
pass misses items, use the two-session workflow instead:

```{r}
ars_copilot_instructions(workflow = "two_phase")
```

which writes a Phase-1 (evidence blueprint) and a Phase-2 (semantic
construction + repair) instruction file. Phase 1 returns
`tlf_extraction_blueprints.json`; you feed that back in session 2 to produce
the supplement.

### Typed conditions (format v4)

The single most important thing to know about the format: **every filter,
population, and column condition is a typed object, not a string.** A row that
displays "Completed" is expressed as

```json
{"rowLabel": "Completed", "variable": {"dataset": "ADSL", "variable": "EOSSTT"},
 "whereClause": {"condition": {"dataset": "ADSL", "variable": "EOSSTT",
                               "comparator": "EQ", "value": ["COMPLETED"]}}}
```

not as `"where": "EOSSTT='COMPLETED'"`. This is what eliminates the old
string-parsing failures (double `=`, smart quotes, `OR` splitting): the
assistant hands the package a parsed tree, so there is nothing left to
mis-parse. The instruction file and the shipped JSON Schema spell the shape out
in full, with a worked example.

### Step 2 — upload to the assistant

Upload **four files together** to your chat assistant:

* `arsbridge_copilot_instructions.md` (from step 1),
* `arsbridge_supplement_v4.schema.json` (so the assistant can self-check),
* your annotated TLF shell `.docx`,
* your ADaM specification `.xlsx`.

Save the assistant's reply — one JSON block — as `supplement.json`.

Consider sending a **draft** as a fifth file. `write_supplement_draft()` runs the
deterministic parse and writes what it found in the same v4 format, one entry
per TLF, with everything it was unsure about listed under
`provenance$reviewItems`:

```{r}
write_supplement_draft(
  shell_path     = "shell.docx",
  adam_spec_path = "adam_spec.xlsx",
  output_path    = "supplement_draft.json"
)
```

The assistant then corrects a structured starting point instead of authoring one
from nothing, and the review items tell it where to look. It is a deterministic
parse — no key, no network, no LLM of any kind — so it costs nothing to produce
before the conversation starts.

### Step 3 — pre-flight (optional but recommended)

```{r}
ars_validate_supplement("supplement.json", "adam_spec.xlsx")
```

This checks the format version, every typed condition, the analysis-type and
method enums, and — when you pass the spec — that every referenced variable
exists in your ADaM spec. Any `FAIL` line is worded so you can paste it back to
the assistant, e.g. *"regenerate: ADSL.WEIGHTBL is not in the ADaM spec"*; when
there are FAILs the result also carries a `repair_prompt` attribute bundling
them into one paste-ready block. When `jsonvalidate` is installed, the file is
additionally checked against the shipped JSON Schema.

### Step 4 — run

```{r}
res <- spec_to_ars(
  shell_path     = "shell.docx",
  adam_spec_path = "adam_spec.xlsx",
  supplement     = "supplement.json"
)
res$extraction_mode
#> "supplement"
```

The supplement has to be named. Nothing is auto-discovered: a `supplement.json`
sitting next to your shell changes nothing until you pass it.

### Step 5 — review the result

The ARS is the deliverable, and it is meant to be read before it is used:

```{r}
ars_workflow()   # the Shiny editor: inspect and correct the converted event
```

## The whole workflow on the bundled study

The training bundle ships a reviewed supplement, so the offline path is
runnable end to end with no key, no network and nothing to author:

```{r}
res <- spec_to_ars_example(supplement = arsbridge_example("supplement.json"))
res$extraction_mode
#> "supplement"
```

It is what a chat assistant could return for the bundled Word shell, corrected
by hand — including the parts a reviewer decides *not* to change, which are
written down in `provenance$reviewItems` rather than left silent. Read it
alongside the shell to see the shape a real reply takes; it is plain JSON:

```{r}
cat(readLines(arsbridge_example("supplement.json")), sep = "\n")
```

Run the same shell without `supplement =` and compare: the enrolled-subjects
analysis set has no condition to filter on, the treatment axis is data-driven
rather than the three columns the shell prints, and the row that heads Table
14.3.2 has no analysis at all. Those are the gaps a Word shell leaves and a
supplement closes.

## What the supplement can and cannot do

The supplement is an **assistant, not an override**. Two guarantees make it
safe to trust a draft it produced:

* **The regex wins (by default).** A supplement binding lands only on a row the
  deterministic pass left unannotated. If the shell already annotated a row,
  the shell's annotation stands and any disagreement is logged as a `WARN`
  for you to review — the assistant can never silently rewrite an authored
  annotation. When you *do* trust the assistant's reading more than the regex
  (a shell the deterministic pass extracts poorly), pass
  `spec_to_ars(supplement_trust = "prefer_supplement")`: a validated,
  spec-gated supplement value then overrides the regex on a conflict, with a
  `WARN` recording both and the shell's original kept as a secondary analysis.
  The spec gate still applies, so a hallucinated variable is rejected in either
  mode.
* **The spec is the gate.** Every `DATASET.VARIABLE` the supplement proposes
  is checked against your ADaM spec. One that is not in the spec is dropped
  and logged as a blocking `FAIL`, exactly as a live LLM's proposal would be.
  The assistant cannot invent a variable into your output.

* **It confirms the table set.** The assistant reads the whole shell
  independently and lists every output with its `title`, so `spec_to_ars()`
  can compare that inventory against what it parsed. A supplement entry with no
  matching table, a table the supplement never mentions, or a title that
  disagrees between the two is logged as a `WARN` — this is the one tier that
  can catch a table arsbridge parsed differently or missed, because the live
  LLM only ever sees sections the parser already found. If the shell heading
  carried no title but the supplement has one, the parsed section adopts it.

Statistic sub-rows (`Mean (SD)`, `Median`, `Q1, Q3`, `Min, Max`) are never
bound from the supplement: they belong to the analysis row above them, and
the layout-driven renderer fills them from that parent.

Everything downstream is unchanged — the ARS JSON is built, validated, and
emitted as per-TLF `{cards}` scripts the same way in every tier. The JSON's
`_meta.extraction_mode` field records which tier produced it, and
`_meta.requires_human_review` stays `TRUE`: whichever tier you used, a
qualified programmer must review the draft.

## Data governance

Uploading your shell and spec to a chat assistant transmits their **text** —
TLF titles, stub labels, footnotes, and the variable *names* from your spec —
to that provider. It does **not** transmit any patient data: shells are mock
by definition, and `spec_to_ars()` never reads ADaM datasets at this stage
(data are only touched later, locally, by `ars_to_ard()`). Even so, shell
wording and a study's variable inventory can be sponsor-confidential —
**confirm your organisation's policy before uploading.**

## Troubleshooting

| Symptom | Cause | Fix |
|---|---|---|
| `Supplement file is not valid JSON` | assistant wrapped the reply in prose, or used smart quotes | the reader strips a ```` ```json ```` fence and smart double-quotes automatically; if it still fails, ask the assistant to "reply with only one fenced strict-JSON block" |
| `Supplement version mismatch` | a v3 file, or one made with a different package version | v4 conditions are typed objects, not strings — regenerate with `ars_copilot_instructions()` from the installed version |
| `invalid whereClause` / `not a where-clause` (`FAIL`) | a condition was written as a string, or with a bad comparator | use the typed shape `{"condition": {"dataset", "variable", "comparator", "value": [...]}}`; paste the `repair_prompt` back to the assistant |
| `analysis label 'X' matched no stub row` (`WARN`) | the `rowLabel` is not the stub text verbatim | copy `rowLabel` exactly from the shell, footnote markers and all |
| `is not in the ADaM spec` (`FAIL`) | assistant proposed a variable your spec does not have | fix the analysis or add the variable to the spec |
| `Anchor mismatch` (`WARN`) | the supplement's first row / row count disagrees with the parsed table | confirm the assistant described the same table arsbridge parsed |
| `entry 'X' matches no TLF in the shell` (`WARN`) | TLF numbering typo in the reply | key each TLF by its shell heading number (e.g. `"14.1.1"`) |
| a TLF ran on heuristics (`WARN`) | that TLF is missing from the supplement | add it to `tlfs` and rerun |
