---
title: "Using arsbridge without API access"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using arsbridge without API access}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE     # Vignette is documentation; calls are illustrative.
)
```

## Three tiers, one required

`arsbridge` has an LLM in the loop for two jobs: reading variable annotations
in shell layouts a regex was never designed for, and classifying each table
(analysis type, grouping columns, method). Neither is required. The reading
engine has three tiers and only the first is mandatory:

| Tier | You supply | Accuracy |
|---|---|---|
| **Deterministic** | shell + spec | Regex + keyword heuristics |
| **Supplement** | + a JSON file from a chat assistant | Near-LLM, no API call |
| **LLM** | + an API key | Full |

A missing key and a missing supplement never stop the run. With neither,
`spec_to_ars()` completes on regex plus heuristics and records the fact:

```{r}
res <- spec_to_ars("shell.docx", "adam_spec.xlsx")
res$extraction_mode
#> "deterministic"
```

Standard shells (variables in red text, `Label -> DATASET.VAR` lines,
bracketed `DATASET.VARIABLE` after a label) still bind correctly with no LLM
at all. What degrades in deterministic mode is variant annotation layouts,
multi-level groupings, the Total column, and analysis-type classification for
unusual tables — each recorded as a `WARN` in the diagnostics sheet.

## The supplement tier: use a chat assistant by hand

Many regulated or enterprise environments block LLM APIs but allow an
interactive chat assistant (GitHub Copilot, ChatGPT, an internal portal)
where you can upload files. The supplement tier turns that into near-LLM
accuracy with zero API calls: you ferry the work through the chat UI, and the
package consumes the assistant's reply as an auxiliary file.

### Step 1 — get the instruction file

```{r}
ars_copilot_instructions()
```

This writes `arsbridge_copilot_instructions.md` into your working directory
and prints the workflow. The file is static and versioned — it tells the
assistant exactly what to read and, critically, the exact JSON shape to
reply with, so the package can validate the answer.

### Step 2 — upload to the assistant

Upload **three files together** to your chat assistant:

* `arsbridge_copilot_instructions.md` (from step 1),
* your annotated TLF shell `.docx`,
* your ADaM specification `.xlsx`.

Save the assistant's reply — one JSON block — as `supplement.json`.

### Step 3 — pre-flight (optional but recommended)

```{r}
ars_validate_supplement("supplement.json", "adam_spec.xlsx")
```

This checks the format version, every binding's shape and variable syntax,
the analysis-type enums, and — when you pass the spec — that every proposed
variable actually exists in your ADaM spec. Any `FAIL` line is worded so you
can paste it straight back to the assistant to regenerate the file, e.g.
*"regenerate: ADSL.WEIGHTBL is not in the ADaM spec"*.

### Step 4 — run

```{r}
res <- spec_to_ars(
  shell_path     = "shell.docx",
  adam_spec_path = "adam_spec.xlsx",
  supplement     = "supplement.json"
)
res$extraction_mode
#> "supplement"
```

## What the supplement can and cannot do

The supplement is an **assistant, not an override**. Two guarantees make it
safe to trust a draft it produced:

* **The regex wins.** A supplement binding lands only on a row the
  deterministic pass left unannotated. If the shell already annotated a row,
  the shell's annotation stands and any disagreement is logged as a `WARN`
  for you to review — the assistant can never silently rewrite an authored
  annotation.
* **The spec is the gate.** Every `DATASET.VARIABLE` the supplement proposes
  is checked against your ADaM spec. One that is not in the spec is dropped
  and logged as a blocking `FAIL`, exactly as a live LLM's proposal would be.
  The assistant cannot invent a variable into your output.

Statistic sub-rows (`Mean (SD)`, `Median`, `Q1, Q3`, `Min, Max`) are never
bound from the supplement: they belong to the analysis row above them, and
the layout-driven renderer fills them from that parent.

Everything downstream is unchanged — the ARS JSON is built, validated, and
emitted as per-TLF `{cards}` scripts the same way in every tier. The JSON's
`_meta.extraction_mode` field records which tier produced it, and
`_meta.requires_human_review` stays `TRUE`: whichever tier you used, a
qualified programmer must review the draft.

## Data governance

Uploading your shell and spec to a chat assistant transmits their **text** —
TLF titles, stub labels, footnotes, and the variable *names* from your spec —
to that provider. It does **not** transmit any patient data: shells are mock
by definition, and `spec_to_ars()` never reads ADaM datasets at this stage
(data are only touched later, locally, by `ars_to_ard()`). Even so, shell
wording and a study's variable inventory can be sponsor-confidential —
**confirm your organisation's policy before uploading.**

## Troubleshooting

| Symptom | Cause | Fix |
|---|---|---|
| `Supplement file is not valid JSON` | assistant wrapped the reply in prose, or used smart quotes | the reader strips a ```` ```json ```` fence and smart double-quotes automatically; if it still fails, ask the assistant to "reply with only one fenced strict-JSON block" |
| `Supplement version mismatch` | file made with a different package version | rerun `ars_copilot_instructions()` from the installed version and regenerate |
| `binding label 'X' matched no stub row` (`WARN`) | the `label` is not the stub text verbatim | ensure `label` is copied exactly from the shell, footnote markers and all |
| `is not in the ADaM spec` (`FAIL`) | assistant proposed a variable your spec does not have | fix the binding or add the variable to the spec |
| `entry 'X' matches no TLF in the shell` (`WARN`) | TLF numbering typo in the reply | key each TLF by its shell heading number (e.g. `"14.1.1"`) |
| a TLF ran on heuristics (`WARN`) | that TLF is missing from the supplement | add it to `tlfs` and rerun |
