---
title: "How arsbridge reads an annotated shell"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How arsbridge reads an annotated shell}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # Documentation; calls are illustrative.
)
```

## The problem: every shell is annotated differently

A TLF shell is a Word document. The lead programmer marks which ADaM
variable feeds each row, but there is **no single convention**. Across
studies, sponsors, and individuals you see the variable reference carried
by:

- a coloured run (`ADSL.AGE` in red),
- bold / italic / underlined text,
- square or round brackets (`[ADAE.AEDECOD WHERE AEREL='RELATED']`),
- plain text appended after the label (`Age (years) ADSL.AGE`),
- or a layout no regex has seen before -- the variable on line 2 of the
  cell, in a sentence, split across runs, mixed-case, abbreviated.

A pure pattern matcher reads the conventions it was written for and
silently misses the rest. arsbridge is built to **extract as many
annotation variants as possible** -- and to never ship a variable it
cannot prove exists.

## The engine: deterministic regex + LLM, together

arsbridge reads each shell **twice**, and takes the union.

```
                 annotated shell (.docx)
                          │
        ┌─────────────────┴─────────────────┐
        ▼                                   ▼
1. DETERMINISTIC PASS               2. LLM PASS (primary)
   parse_shell_docx()                  extract_shell_llm()
   4-layer regex detector              re-reads each raw cell,
   • colour                            separates label vs variable
   • bold/italic/underline             in ANY layout
   • plain-text DATASET.VAR
   • (gap → handed to LLM)
        │                                   │
        └─────────────────┬─────────────────┘
                          ▼
            3. HARD ADaM-SPEC GATE
       every proposed DATASET.VARIABLE must
       exist in the ADaM spec, or it is
       REJECTED and logged as a blocker
                          ▼
              validated annotations → ARS JSON
```

### Pass 1 -- deterministic regex (fast, free, offline)

`parse_shell_docx()` walks the document's OOXML and runs a four-layer
detector on every stub cell and listing header. It recognises the known
conventions above with no API call. This pass alone fully handles shells
that follow a standard annotation style, and it always runs -- even with
no API key.

### Pass 2 -- the LLM as primary reader (handles the variants)

`extract_shell_llm()` re-reads the **raw** text of each cell and asks the
model to separate the human display label from the machine variable
reference, in whatever layout the shell happens to use. This is where the
"as many variants as possible" power comes from: the LLM generalises to
formats the regex was never written for. It returns structured output
(via an `ellmer` type), never free text to parse.

### Pass 3 -- the hard spec gate (the seatbelt)

The LLM is powerful but can hallucinate. So **every** proposed
`DATASET.VARIABLE` is checked against your ADaM specification
(`define.xml` or Excel). A proposal that is not in the spec is **dropped,
not shipped**, and recorded as a blocking finding that names the row and
the rejected token. The spec is the ground-truth oracle: the model can
only pick variables that actually exist in your study.

This preserves arsbridge's founding promise -- *it extracts and
converts; it does not invent* -- while letting the LLM read freely. The
promise is now **enforced by the gate**, not by limiting the reader.

## How the two passes combine (coverage table)

A row's annotation is taken if **either** pass finds it. On a conflict,
the LLM wins (it is the primary reader) and a warning flags the
disagreement so a human can check it.

| Row situation                         | Regex | LLM (in spec) | Result                         |
|---------------------------------------|:-----:|:-------------:|--------------------------------|
| Known convention                      |  ✓    | silent        | **regex kept**                 |
| Known convention                      |  ✓    | ✓ same        | confirmed                      |
| Known convention, different reads     |  ✓    | ✓ different   | **LLM wins** + `WARN`          |
| **Variant layout regex can't parse**  |  ✗    | ✓             | **LLM adds it** (the win)      |
| LLM proposes a non-spec variable      |  ✗    | ✗ (rejected)  | dropped + **blocker**          |
| Genuinely no variable                 |  ✗    | silent        | empty                          |

Guards keep the regex catch safe: if the LLM returns an empty
dataset/variable, omits a row, or no API key is set, the deterministic
result stands.

## Degraded mode: it still runs with no key

No API key, or `extract_with_llm = FALSE`, means Pass 2 is skipped and
arsbridge runs on the deterministic regex alone, emitting one warning
that variant-format extraction is unavailable. The pipeline still
produces ARS, ARD, and code for standard shells -- offline and in CI.

```{r}
res <- arsbridge::spec_to_ars(
  shell_path     = "inputs/shells.docx",
  adam_spec_path = "inputs/adam_spec.xlsx",
  extract_with_llm = TRUE     # default; FALSE = regex only
)
```

## Every gap is reported in plain English

Nothing is dropped silently. Rejected variables, regex/LLM
disagreements, LLM outages, and unparsed rows all land in the diagnostics
and blockers, with the **input document to fix** named:

```{r}
arsbridge::ars_diagnostics(res$diagnostics)  # everything recorded
arsbridge::ars_blockers(res$diagnostics)     # only what blocks clean output
```

A blocker tells you *what* (rejected variable / missed row), *why*
(not in spec / unknown convention), and *how to fix* (correct the shell
annotation, or add the variable to the ADaM spec).

## Choosing and adding a model

The reader works with any provider in the registry -- Anthropic, OpenAI,
Gemini, or an OpenAI-compatible one such as GLM. For clinical text,
Anthropic has the lowest false-positive content-filter rate; some
providers block adverse-event terminology.

```{r}
arsbridge::set_anthropic_key()                 # recommended for clinical text
arsbridge::set_llm_key("glm", "your-glm-key")  # any registry provider
Sys.setenv(ARS_LLM_PROVIDER = "anthropic")     # pick the active one
arsbridge::show_active_llm()
```

Adding a brand-new provider is a single entry in the provider registry
(`R/llm_providers.R`): its key environment variable, default model, the
`ellmer` chat constructor, and a `base_url` if it is OpenAI-compatible.
No other code changes.
